Tidal Heating via Internal Gravity Waves in White Dwarf Binaries

### 2018-12-30 writeups/paper.tex
stash

    \documentclass[twocolumn,
        usenames, % allows access to some tikz colors
        dvipsnames % more colors: https://en.wikibooks.org/wiki/LaTeX/Colors
    ]{revtex4-1}% chktex 8
    \usepackage{
        amsmath,
        amssymb,
        fancyhdr, % page styling
        lastpage, % footer fanciness
        hyperref, % various links
        setspace, % line spacing
        amsthm, % newtheorem and proof environment
        mathtools, % \Aboxed for boxing inside aligns, among others
        float, % Allow [H] figure env alignment
        enumerate, % Allow custom enumerate numbering
        graphicx, % allow includegraphics with more filetypes
        wasysym, % \smiley!
        upgreek, % \upmu for \mum macro
        listings, % writing TrueType fonts and including code prettily
        tikz, % drawing things
        booktabs, % \bottomrule instead of hline apparently
        cancel % can cancel things out!
    }
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[font=scriptsize]{subcaption} % subfigures

@@

\title{IGWs and WDs Ongoing Writeup}
\author{Yubo Su}
\maketitle

@@

WIP ongoing writeup, last updated \today.

% Define white dwarfs (WD), internal gravity waves (IGW), citations

Basic narrative is: the prevailing understanding is that mean flow absorption
slowly spins up a WD from the outside in as the critical layer propagates
inwards. Our numerical simulations validate this prediction and give an
analytical prediction for the motion of the critical layer. However, it appears
our numerical simulations are viscosity limited. We present a brief argument for
why the viscosity limitations are insignificant, but also present some
exploratory, inconclusive work towards direct numerical simulation resolving
critical layer behavior without being viscosity limited.

In \autoref{s:eq}, we will cover the equations and numerical setup. In
\autoref{s:lin}, we will discuss the agreement of our simulation in the
low-amplitude limit with analytical linear theory. In \autoref{s:nonlin} we will
present a few reference simulations illustrating critical layer absorption
depositing horizontal angular momentum into the fluid. Finally, in
\autoref{s:bouss} we will consider a speculative series of simulations
attempting to rectify the seemingly viscosity-limited simulations of the
previous section.

\section{Equations}\label{s:eq}

We simulate an incompressible, isothermal fluid, representatitive of degenerate
matter in WD bulks. We assume a barotropic equation of state as a first
approximation. Furthermore, we model the background, hydrostatic density
stratification as $\rho_0(x, z) = \rho_0(z) \propto e^{-z/H}$.

The physical equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:fc_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\rho}{t} + \vec{u} \cdot \vec{\nabla}\rho &= 0
            ,\label{eq:fc_density}\\
        \pd{\vec{u}}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\vec{u}
            + \frac{\vec{\nabla}P}{\rho}
            + g\hat{z} &= 0.
    \end{align}
\end{subequations}
$\vec{u}, \rho, P$ are the fluid dynamical variables and $-g\hat{z}$ is constant
gravitational acceleration. Note that at hydrostatic equilibrium $\partial_t =
0$ we have $\vec{u} = 0, \vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 =
\rho_0 gH$.

We nondimensionalize by taking $H = N = \rho_{ref} = 1$ where
\begin{equation}
    N^2 \equiv g^2\p*{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
is the Brunt-V\"ais\"al\"a frequency and $\rho_{ref}$ is the background density
stratification at $z = 0$. The full numerical details of the simulation are
described in \autoref{ss:strat_impl}.

@@

To model a continuous IGW wave train excited deep in the WD interior propagating
towards the surface without unnecessary computational cost, we excite an
IGW wave train from the bottom of the simulation domain. As Chebyshev grid is
a factor of $\propto N_z$ denser near the edges of the simulation domain, we
use a volumetric forcing to excite IGW at some $z_0 > 0$ interior to the
simulation domain. Our forcing excites both IGW propagating upwards, the
physically interesting solution, and downwards, which are damped away by damping
layers (see \autoref{ss:strat_impl}) before inducing strict Courant-Friedrichs-Lewy
(CFL) timestepping constraints (cite).

As not to interfere with the incompressibility constraint, we force the system
on the density equation, replacing \autoref{eq:fc_density} with
\begin{equation}
    \pd{\rho}{t} + \vec{u} \cdot \vec{\nabla}\rho =
        Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p*{k_{0, x}x - \omega_0 t}.
\end{equation}
Define $k_{0, z}: \omega(k_{0, x}, k_{0, z}) = \omega_0$. If $k_{0, z}\sigma \ll
1$, the forcing term has the effect of picking out $k_{0, z}$ driven by
$\cos\p*{k_{0, x}x - \omega_0 t}$ from the broad $k_z$ wavenumber spectrum
permitted by the narrow Gaussian profile.

We first test our numerical simulation with a ``linear'' simulation, where
all flow quantities are small: we assume  $\rho_1 \equiv \rho - \rho_0 \ll
\rho_0$ and $\vec{u} \cdot \vec{\nabla} \ll \partial_t$. We clarify that our
``linear'' simulations solve the full nonlinear equations, just in the weak
perturbation limit.

\subsection{Analytical Solution}

In the small perturbation limit, we linearize \autoref{se:fc_orig} with
inhomegeneous driving term and obtain:
\begin{subequations}\label{se:vol_drive}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}_1}{t} + \frac{\vec{\nabla}P}{\rho_0}
            + \frac{\rho_1 \vec{g}}{\rho_0}
            &= 0,\\
        \pd{\rho_1}{t} - \frac{u_{1z} \rho_0}{H}
            &= Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
                e^{i(k_xx - \omega t)}.
    \end{align}
\end{subequations}
We have substituted $\cos\p*{k_xx - \omega_0t}$ for a complex exponential for
easier manipulation; taking real parts of all dynamical variables above recovers
the physical solution.

@@

To solve the inhomogeneous system, we may first approximate the driving term
using $e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$ if $\sigma \to 0$, then relax to $k_{0z}\sigma \lesssim 1$. This may be
solved to obtain the solution
\begin{align}
    A(F) &= \frac{Fgk_{0x}^2}{\rho_0(z_0)\omega_0^2}
        \frac{1}{2ik_z}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}},\\
    u_{1z} &= A(F) \times\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i(k_{0x}x + k_{0z}(z - z_0) - \omega_0 t)}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i(k_{0x}x - k_{0z}(z - z_0) - \omega_0 t)}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The extra factor of $e^{-\frac{(k_{0z}\sigma)^2}{2}}$ compared to the
$\delta$-function solution comes from evaluating the Fourier Transform of
$e^{-\frac{(z - z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$.

@@

@@

\section{Boussinesq Local Simulation (Speculative)}\label{s:bouss}

Alternatively, we can attempt direct numerical simulation probing lower
Richardson numbers. We will move to the Boussinesq approximation and set up a
mean flow as an initial condition. This relaxes the constraint that the domain
must be sufficiently large to permit substantial $e^{z/H}$ growth of the mean
flow $\bar{U}_0$ and allows us to probe much finer vertical resolution.

The physical equations in the Boussinesq approximation are:
\begin{subequations}\label{se:bouss}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\rho_1}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\rho_1
            - N^2\frac{\rho_0 u_z}{g} &= 0,\\
        \pd{\vec{u}}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\vec{u}
            + \frac{\vec{\nabla}P_1}{\rho_0}
            + \frac{\rho_1 g\hat{z}}{\rho_0} &= 0.
    \end{align}
\end{subequations}
In these equations, we take $\rho_0(z) = \rho_0 = 1$ a constant reference
density. We nondimensionalize $\rho_0 = N = g = 1$. By using a subgrid model
(details \autoref{ss:bouss_impl}), we are able to probe significantly smaller
Richardson numbers.

\subsection{Low-Resolution Simulation}

The parameters of the simulation differ slightly; since we do not have
exponential density stratification to induce $\bar{U}_0 \propto e^{z/H}$ growth,
we initialize the system with an initial $\bar{U}_0(z) \leq \bar{U}_c$,
simulating a later phase of the stratified simulation where the critical layer
has propagated deep into the domain.

Sample snapshots from a reference simulation are \autoref{fig:bouss_1}.
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/vstrat_1.png}
        \caption{Boussinesq simulation initial condition.}
    \end{subfigure}

    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/vstrat_2.png}
        \caption{At later time, when critical layer absorption has begun.}
    \end{subfigure}

    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/vstrat_3.png}
        \caption{At an even later time, a strong reflective character seems to
        emerge as the Richardson number $\to 0.5$.}
    \end{subfigure}
    \caption{Boussinesq simulation snapshots.}\label{fig:bouss_1}
\end{figure}
The resulting front velocity and Richardson number are depicted in
\autoref{fig:bouss}. Most important to note are the significantly diminished
critical layer velocity compared to the prediction (owing to critical layer
reflection) and the Richardson number which approaches the critical $1/4$ value
for KHI-instability.
\begin{figure}[t]
    \centering
    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/vstrat_ri.png}
        \caption{Richardson number evolution over time.}
    \end{subfigure}

    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/vstrat_front_v.png}
        \caption{Front velocity compared to analytical \autoref{eq:u0_lin}
        prediction.}
    \end{subfigure}
    \caption{Boussinesq data analysis plots.}\label{fig:bouss}
\end{figure}

YUBONOTE\@: The next step would be to run simulations at sufficient resolution to
resolve the KHI (only possible in the Boussinesq system, where we are not
required to have a sufficiently large vertical domain to capture exponential
growth and can choose ever smaller vertical domains); I have started to try this
but haven't resolved the resultant numerical issues yet.

@@

\subsection{Boussinesq Fluid}\label{ss:bouss_impl}

In the Boussinesq system, there is no need to transform to $T, \Upsilon$
variables as $\rho_0$ is constant. Including numerical terms and driving terms,
our full equations are
\begin{subequations}\label{se:bouss_num}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\rho_1}{t} - \frac{\rho_0 u_z}{H}
            - \nu \nabla^6 \rho_1
            &= -\Gamma(z) \rho_1
                - \p*{\vec{u} \cdot \vec{\nabla}}\rho_1
                + Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
                    \cos \p*{k_xx - \omega t},\\
        \pd{u_x}{t} + \frac{\partial_x P}{\rho_0}
            - \nu \nabla^6 u_x
            &= -\Gamma(z) u_x
                - \p*{\vec{u} \cdot \vec{\nabla}}u_x,\\
        \pd{u_z}{t} + \frac{\partial_z P}{\rho_0}
            + \frac{\rho_1 g}{\rho_0}
            - \nu \nabla^6 u_z
            &= -\Gamma(z) u_z
                - \p*{\vec{u} \cdot \vec{\nabla}}u_z.
    \end{align}
\end{subequations}
$\Gamma(z)$ is as \autoref{eq:Gamma}. We use subgrid model $\nabla^6$ (cite) to
decrease the effect of viscosity on scales above the grid resolution.

>>>>

    \documentclass[twocolumn,
        nofootinbib,
        usenames, % allows access to some tikz colors
        dvipsnames % more colors: https://en.wikibooks.org/wiki/LaTeX/Colors
    ]{revtex4-1}% chktex 8
    \usepackage{
        amsmath,
        amssymb,
        fancyhdr, % page styling
        lastpage, % footer fanciness
        hyperref, % various links
        setspace, % line spacing
        amsthm, % newtheorem and proof environment
        mathtools, % \Aboxed for boxing inside aligns, among others
        float, % Allow [H] figure env alignment
        enumerate, % Allow custom enumerate numbering
        graphicx, % allow includegraphics with more filetypes
        wasysym, % \smiley!
        upgreek, % \upmu for \mum macro
        listings, % writing TrueType fonts and including code prettily
        tikz, % drawing things
        booktabs, % \bottomrule instead of hline apparently
        cancel % can cancel things out!
    }
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[font=scriptsize]{subcaption} % subfigures

@@

\begin{abstract}
    In white dwarf binaries tidal torques are expected to raise a train of
    internal gravity waves in the white dwarf interior. These
    outward-propagating waves can then grow to be nonlinear and break via
    hydrodynamical instabilities, transfering energy and angular momentum from
    the binary orbit to the white dwarf. We perform 2D numerical simulations of
    nonlinear wave breaking of outwards-propagating internal gravity waves in an
    incompressible, isothermal atmosphere with an exponentially decaying density
    stratification. We find that tidal synchronization, after an initial
    transient phase, proceeds inwards from the surface in thin layers. We argue
    the thickness of these layers is limited by the Kelvin-Helmholtz
    Instability. We provide simple analytical formulae for the location and
    thickness of tidal synchronization over time that are in good agreement with
    our simulations.
\end{abstract}

\title{Tidal Heating via Internal Gravity Waves in White Dwarf Binaries}
\author{Authors}
\date{\today}
\maketitle

@@

% Usual WD binaries introduction, from proposals

% IGW tidal dissipation has been thoroughly studied in many astrophysical
% binary systems (Barker Ogilvie, Goldreich Nicholson, Zahn?). Its effect in WDs
% (Fuller Lai) resembles early solar type stars but has not been studied
% numerically. While the linear IGW solution is an exact solution of the
% nonlinear system (Drazin) but is globally unstable (Drazin), the IVP must be
% studied numerically to determine how the linear solution breaks down and
% deposits energy/angular momentum.

% From numerical studies, IGWs are known to be anti-diffusive (Lecoanet mean
% flow papers, Lindzen) and sharpen mean flows (m = 0). Mean flows produce
% reflection, known both analytically (Booker & Bretherton) and numerically
% (Winters d'Asaro). IGW self-interaction via its wave-induced mean flow is
% known to play a crucial role in IGW wavebreaking (Sutherland x2).

% We study IGW breaking in the plane-parallel approximation in 2D. Known to
% describe both wave steepening (Kloyermeister 1991) and captures mean flow
% absorption (Booker & Bretherton). While the details of breaking are presently
% understood to be 3D (Winters & d'Asaro), we argue in (section) that our
% results should be qualitatively similar to 3D (both 2D and 3D exhibit KHI, and
% since we are able to resolve KHI, the spinup in 3D should be similar?).

In \autoref{s:num_setup}, we will cover the numerical setup. In \autoref{s:lin}, we
will discuss the agreement of our simulation in the low-amplitude limit with
analytical linear theory. In \autoref{s:nonlin} we will present a few reference
simulations illustrating critical layer absorption depositing horizontal angular
momentum into the fluid.

\section{Problem Description}\label{s:num_setup}

\subsection{Fluid Equations}

We consider an incompressible, isothermal fluid, representative of degenerate
matter in WD bulks, in 2D. We assume a barotropic equation of state as a first
approximation. As we are interested in the behavior of the wave far from the
center of the WD, we approximate the gravitational field as uniform. We model
the background density stratification as $\rho_0(x, z) = \rho_0(z) = \rho_{ref}
e^{-z/H}$ for some reference density $\rho_{ref} = \at{\rho_0(z)}_{z = 0}$.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:fc_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \md{\rho}{t} &= 0 ,\label{eq:fc_density}\\
        \md{\vec{u}}{t} + \frac{\vec{\nabla}P}{\rho} + g\hat{z} &= 0.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p*{\vec{u} \cdot \vec{\nabla}}$ is the material
derivative. $\vec{u}, \rho, P$ are the fluid dynamical variables and $-g\hat{z}$
is constant gravitational acceleration. Note that at hydrostatic equilibrium
$\pd{}{t} = 0$ we have $\vec{u} = 0, \vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and
so $P_0 = \rho_0 gH$.

We nondimensionalize by taking $H = N = \rho_{ref} = 1$ where
\begin{equation}
    N^2 \equiv g^2\p*{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
is the Brunt-V\"ais\"al\"a frequency. The full numerical details of the
simulation are described in \autoref{ss:strat_impl}.

@@

To model a continuous IGW wave train excited deep in the WD interior propagating
towards the surface without unnecessary computational cost, we excite an
IGW wave train from the bottom of the simulation domain. We use a volumetric
forcing term to excite IGW at some $z_0 > 0$ interior to the simulation
domain\footnote{Interfacial forcing at the bottom boundary incurs severe CFL
timestepping constraints, as we use a Chebyshev polynomial basis along the $z$
axis in our spectral method which has very small grid spacing at the
boundaries.}. Our forcing excites both IGWs propagating upwards, imitating a
wave tidally excited deeper in the WD, and downwards, which are damped away by
damping layers (see \autoref{ss:numerics}) before inducing strict
Courant-Friedrichs-Lewy (CFL) timestepping constraints.

As not to interfere with the incompressibility constraint, we force the system
on the density equation, replacing \autoref{eq:fc_density} with
\begin{equation}
    \md{\rho}{t} = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p*{k_{0x}x - \omega_0 t}.
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{0z}$ satisfying the IGW's dispersion relation for the given
$k_{0x}, \omega_0$ will propagate.
% Define $k_{0z}: \omega(k_{0x}, k_{0z}) = \omega_0$ the wavenumber picked
% out by $k_{0x}, \omega_0$. If $k_{0z}\sigma \ll 1$, the $\cos\p*{k_{0x}x -
% \omega_0 t}$ term permits only the $k_{0z}$ mode to propagate among the broad
% $k_z$ wavenumber spectrum permitted by the narrow Gaussian profile.

\subsection{Internal Gravity Waves}

In the small perturbation limit, we linearize \autoref{se:fc_orig} with
inhomegeneous driving term and obtain:
\begin{subequations}\label{se:vol_drive}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}_1}{t} + \frac{\vec{\nabla}P}{\rho_0}
            + \frac{\rho_1 \vec{g}}{\rho_0}
            &= 0,\\
        \pd{\rho_1}{t} - \frac{u_{1z} \rho_0}{H}
            &= Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
                e^{i(k_xx - \omega t)}.
    \end{align}
\end{subequations}
We have substituted $\cos\p*{k_xx - \omega_0t}$ for a complex exponential for
easier manipulation.

@@

To solve the inhomogeneous system, we may first approximate the driving term
using $e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$ if $\sigma \to 0$. This system is solved exactly by matching the two
homogeneous solutions above and below $z_0$, We may then approximately relax the
solution to $k_{0z}\sigma \lesssim 1$. This may be solved to obtain the solution
\begin{align}
    A(F) &= \frac{Fgk_{0x}^2}{\rho_0(z_0)\omega_0^2}
        \frac{1}{2ik_z}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}},\\
    u_{1z} &= A(F) \times\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i(k_{0x}x + k_{0z}(z - z_0) - \omega_0 t)}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i(k_{0x}x - k_{0z}(z - z_0) - \omega_0 t)}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
An extra factor of $e^{-\frac{(k_{0z}\sigma)^2}{2}}$ arises compared to the
$\delta$-function solution when evaluating the Fourier Transform of
$e^{-\frac{(z - z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$.

% Linear solution: e^z/2H, dispersion relation, phase/group velocity. Observe is
% exact solution of nonlinear.
% Linear solution is globally unstable to RTI though, thus not stable
% equilibrium of IVP
%
% Nonlinear: Critical layer absorption (viscous = Hazel, nonlin = Winters),
% wave-induced mean flow, horizontal momentum flux

@@

\subsection{Numerical Parameters}\label{ss:numerics}

% Description of spectral method, basis functions, initial conditions
% Grid Size
% Viscosity + diffusivity, damping layers (to simulate radiative BC)
% timestepping? RK443, (CFL limiter??)

We first test our numerical simulation with a ``linear'' simulation, where
all flow quantities are small: we assume  $\rho_1 \equiv \rho - \rho_0 \ll
\rho_0$ and $\vec{u} \cdot \vec{\nabla} \ll \partial_t$. We emphasize that our
``linear'' simulations solve the full nonlinear equations, just in the weak
perturbation limit.

@@

@@
